
# Transformer Implementation {#implementation}

There are a numerous blogs/tutorials demonstrating transformer implementations in TensorFlow from scratch, including the [official 
TensorFlow transformer tutorial](https://www.tensorflow.org/tutorials/text/transformer "Transformer model for language understanding").
However, we could not find a single example using the high-level Keras API for building and training the transformer. For example, the official tutorial does not use Keras' built-in APIs for training and validation.
This created difficulties for us when we attempted to build a customized transformer based on existing examples.

The purpose of this article is to present a TensorFlow implementation of the transformer sequence-to-sequence architecture @attention in Keras following the high-level API specifications. We use TensorFlow's built in implementation of the Keras API (see e.g. [Guidance on High-level APIs in TensorFlow 2.0](https://blog.tensorflow.org/2018/12/standardizing-on-keras-guidance.html "Standardizing on Keras: Guidance on High-level APIs in TensorFlow 2.0")).
Using a high-level API makes the learning process more straightforward and the code much briefer. It also avoids reinventing the wheel which can potentially introduce errors. 

While the primary emphasis is on implementation, we also give our own in depth explanation of the transformer model. 

## Requirements

This library requires TensorFlow version 2.5.0. It *may* work on newer versions as well, and we have tested it on the version 2.6 development branch. The full requirements are listed in `inst/python/requirements.txt`. 

Please note the root directory for the actual `python` code is `inst/python`. 

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# FIX THIS!!!
python_working_dir <- here::here('../inst/python')
knitr::opts_knit$set(root.dir = python_working_dir)
knitr::opts_knit$set(width = 80)
library(reticulate)
PATH_v <- Sys.getenv('PATH')
HOME_v <- Sys.getenv('HOME')
Sys.setenv(PATH = paste(PATH_v, paste(HOME_v, '.local/bin', sep = '/'), sep = ':'))
system("pip install pipenv")
system("pipenv --python 3.8")
venv <- system("pipenv --venv", inter = TRUE)
use_virtualenv(venv, required = TRUE)
py_config()
```

```{python requirements, include=FALSE}
import os

os.system("pip install -r requirements.txt")
```

```{r formatting-setup,echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}
# Set so that long lines in R will be wrapped:
#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy=TRUE)
```

```{r test, root.dir=python_working_dir, include=FALSE}
#os <- import('os')
#os$getcwd()
```
